---
output: html_document
editor_options: 
  chunk_output_type: inline
---
# Categorical data

## Aesthetics: group

```{r}
ggplot() +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3),
      y=c(2, 3, 2),
    )
  ) +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3),
      y=c(5, 2, 6)
    )
  )

```

```{r}
ggplot() +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(2, 3, 2, 5, 2, 6),
      group=c("m", "m", "m", "f", "f", "f")
    )
  )
```


Use `group` aesthetic to combine 
  
  * multiple same geom layers

into one.

```{r}
ggplot() +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(2, 3, 2, 5, 2, 6),
      group=c("m", "m", "m", "f", "f", "f") 
    )
  )

```

```{r}
ggplot() +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(2, 3, 2, 5, 2, 6),
      group=c("m", "m", "m", "f", "f", "f"),
      color=c("m", "m", "m", "f", "f", "f")
    )
  )

```

  * Any aesthetic differentiates group can replace group.
  
```{r}
ggplot() +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(2, 3, 2, 5, 2, 6),
      # group=c("m", "m", "m", "f", "f", "f"),
      color=c("m", "m", "m", "f", "f", "f")
    )
  )

```

<div class="alert alert-warning">

  * When there is no aesthetic mapping to differentiate groups, use `group` aesthetic mapping.
  
</div>

## Geom overlapping

When geom layers overlap, we can use

  * `alpha` aesthetic. 

If multiple geometries are created within the one `geom_` call (using grouping aesthetics), we can also set

  * position: "stack", "dodge" or "jitter" (some of them might not apply to certain `geom_`)

```{r , eval=T, echo=F}
teachDS::img_centering("https://clauswilke.com/dataviz/visualizing_proportions_files/figure-html/health-vs-age-1.png")
```

  * <https://clauswilke.com/dataviz/visualizing-proportions.html#fig:health-vs-age>
  
  * x is continuous, or discrete with many types
  
  * y the cumulative proportion

***

```{r}
ggplot() +
  geom_area(
    mapping=aes(
      x=c(1, 2, 3),
      y=c(0.2, 0.3, 0.2), # cat 1 proportion
    )
  ) +
  geom_area(
    mapping=aes(
      x=c(1, 2, 3),
      y=c(0.4, 0.3, 0.52) #+ c(0.2, 0.3, 0.2) # the additive is for accumulative purpose # cat2
    )
  )
```

## Position: stack

  * put y on top of the overlapping geom's y
  
  * create accumulative result.

```{r}
ggplot() +
  geom_area(
    mapping=aes(
      x=c(1, 2, 3, 
        1, 2, 3),
      y=c(0.2, 0.3, 0.2, 
        0.4, 0.3, 0.52),
      # group=c("m", "m", "m", 
        # "f", "f", "f"),
      fill=c("m", "m", "m", 
        "f", "f", "f")
    ),
    position="stack" #input$position
  )
```

  * stack position is accumulative; no need to compute the accumulative value yourself.
  
  * the default position in `geom_area` is "stack". Therefore, you can omit position argument. 


***

```{r data_cat1}
data_cat1 <- data.frame(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(0.2, 0.3, 0.2, 0.4, 0.3, 0.52),
      fill=c("m", "m", "m", "f", "f", "f")
)
```

```{r}
ggplot(
  data=data_cat1
) + 
  geom_area(
    mapping=aes(
      x=x,
      y=y,
      fill=fill
    )
  )
```

```{r}
class(data_cat1$fill)
data_cat1$fill |> class() # same as dplyr %>%
```
```{r}
data_cat1$fill |>
  factor() -> fill_asFactor
fill_asFactor
levels(fill_asFactor)
```


<div class="alert alert-warning">
When aesthetic mapping involves with unordered data, it will  

  * convert the data series into factor (unless the series is already a factor);
  
  * conduct the mapping according to the level sequence of the converted factor.

</div>

```{r}
data_cat1$fill |>
  factor() |>
  levels()
```


## Factor

When grouping aesthetics vary the look of geometries across different groups of data, it is crucial that users declare the mapped series with proper class.

  * `factor(data_series, levels)` parses `data_series` into a categorical data with expressing sequence defined by `levels`.
  
  * If omit `levels` the level sequence will be determined by the collateral sequence defined by your operating system.

```{r}
ggplot(
  data=data_cat1
) + 
  geom_area(
    mapping=aes(
      x=x,
      y=y,
      group=factor(fill, levels=c("f","m")), # stacking sequence is determined by group
      fill=factor(fill, levels=c("m", "f")) # parsing
    )
  )
```

  * Here we declare factor on-the-go.
  
We can also declare factor in the data frame first:
```{r}
data_cat1_copy <- data_cat1
data_cat1_copy$fill |>
  factor(levels=c("m", "f")) -> 
  data_cat1_copy$fill
```

  * `|>` is a R 4.0+ equipped operator, which makes:
```{r}
f(x, ....) # equivalent to
x |> f(...)
```
  

```{r}
ggplot(
  data=data_cat1_copy
) + 
  geom_area(
    mapping=aes(
      x=x,
      y=y,
      group=factor(as.character(fill), levels=c("f", "m")),
      fill=fill
    )
  )

```

## Proportional data

```{r}
data_cat2_wide <- data.frame(
      x=c(1, 2, 3),
      y_a=c(0.2, 0.3, 0.2),
      y_b=c(0.4, 0.4, 0.52),
      y_c=c(0.4, 0.3, 0.28)
)

data.frame(
  x=c(1, 2, 3, 1, 2, 3, 1, 2, 3),
  y =c(0.2, 0.3, 0.2, 0.4, 0.4, 0.52, 0.4, 0.3, 0.28), # values_to
  fill =c("y_a", "y_a", "y_a", "y_b", "y_b", "y_b", "y_c", "y_c", "y_c") # names_to
)

data_cat2_wide |> 
  tidyr::pivot_longer(
    cols=y_a:y_c,
    names_to = "fill",
    values_to= "y"
  ) ->
  data_cat2

View(data_cat2)
```

```{r}
data_cat2$fill |> class()
```


```{r}
ggplot(
  data=data_cat2
) + 
  geom_area(
    mapping=aes(
      x=x,
      y=y,
      group=factor(fill, levels=c("y_c", "y_a", "y_b")),
      fill=factor(fill, levels=c("y_a", "y_b", "y_c"))
    ),
    # position="stack" ,
    color="white" # 邊線色
  )
```

***

When x mapping series has limited cases and is discrete, a bar chart with position dodge is better.


```{r}
ggplot(
  data=data_cat2
) + 
  geom_col(
    mapping=aes(
      x=x,
      y=y,
      group=factor(fill, levels=c("y_c", "y_a", "y_b")),
      fill=factor(fill, levels=c("y_a", "y_b", "y_c"))
    ),
    color="white",
    width=2, #input$width
    size=1, #input$size
    position = "dodge" #input$position
  )
```

  * `width`: the width of the bar
  
  * `size`: the size of the stroke
  

***
Pie chart: 

  * not good for comparing proportion across more than one dimension 

```{r}
library(dplyr)
data_cat2 %>%
  filter(
    x==1
  ) -> 
  data_cat2_x1only
```

```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  )
```

```{r piechart}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  coord_polar(
    theta = "y"
  )
```

***

adding text

```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=fill
    ),
    position = "stack"
  )
```

  * `geom_col` stack sequence is based on `fill` level sequence.
  
  * `geom_text` stack sequence is based on observation sequence.  
  
    * we can use `group` aesthetic to ask `geom_text` to draw group by group.
    
```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=fill,
      group=fill
    ),
    position = "stack"
  )
```

Change label to the proportion data `y`
```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=y,
      group=fill
    ),
    position = "stack"
  )
```

  * `position` argument also takes position functions.
  
  * When you know what type of position you want, you can use corresponding position function to finetune the position.

```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=y,
      group=fill
    ),
    position = position_stack(vjust=0.5)
  )
```

```{r piechart2}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=y,
      group=fill
    ),
    position = position_stack(vjust=0.5)
  ) +
  coord_polar(
    theta = "y"
  ) +
  theme_void()
```


## More on position

<https://ggplot2.tidyverse.org/reference/index.html#section-position-adjustment>

## Exercise

### 1{-}

```{r, echo=F, eval=T}
teachDS::img_centering("https://images.squarespace-cdn.com/content/v1/55b6a6dce4b089e11621d3ed/1586289284165-6Z2M2XKD645EJRCV3TPV/Area+chart+example.png?format=1500w")
```


```{r}
set.seed(2020)
type1 <- function(){
  dnorm(1990:2010, mean=1991, sd=3)*2500 -> x
    round(x, digits = 0)
}
type2 <- function(){
  dnorm(1994:2010, mean=1998, sd=2)*1000->x
  round(x, digits=0)
}
type3 <- function(){
  dnorm(2002:2010, mean=2005, sd=2.3)*800->x
  round(x, digits=0)
}

data_set4 <- data.frame(
  year=c(
    1990:2010,
    1994:2010,
    2002:2010),
  storage_type=
    c(
      rep("type1", 21),
      rep("type2", 17),
      rep("type3", 9)),
  sales_amount=c(
    type1(), 
    type2(),
    type3()
  )
)

```

