# Scale

In geom, `aes(xxx=yyy)`, defines a mapping from data vector `yyy` to aesthetic vector `xxx`. To change or fine tune this mapping, you add `scale_xxx_...` to your ggplot object.

In general, a mapping can be divided into two parts

  * How to map?  
    * `limits`: the range of data vector to be mapped.  
    * `values`: the mapped values from `limits`.  
    
  * In the the mapping, what to show to the readers? This is about the information in **legend** (圖例說明)  
    * `breaks`: the picked values in `limits`.  
    * `labels`: the mapped values of `breaks`.  
    * `name`: the title of the legend.

Depending on aesthetics, not all five settings will be necessary. 

## Legend

```{r}
data_cat1 <- data.frame(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(0.2, 0.3, 0.2, 0.4, 0.4, 0.52),
      fill=c("m", "m", "m", "f", "f", "f")
)
ggplot0 <- list()
0 #input$null
ggplot0$plot1 <- 
  ggplot(
    data=data_cat1
  ) + 
    geom_area(
      mapping=aes(
        x=x,
        y=y,
        fill=factor(fill, levels=c("m", "f"))
      )
    )
ggplot0$plot1
```

```{r}
labels = as.character(seq(1.0, 3.0, by=0.5))
labels
breaks = seq(1.0, 3.0, by=0.5)
breaks 
ggplot0$plot1 +
  scale_x_continuous(
    labels=labels,
    breaks=breaks
  )
```

```{r}
# fill 
labels= c("男", "女")
breaks= c("m", "f")
ggplot0$plot1 +
  scale_fill_discrete(
    name="性別",
    labels=labels, 
    breaks=breaks
  )
```



```{r}
ggplot0$plot1 +
  scale_fill_discrete(
    name="Gender",
    breaks=c("m", "f"),
    labels=c("Male", "Female")
  ) -> ggplot0$plot2
ggplot0$plot2
```


<div class="alert alert-info">
On every graph, by default, there is legend and/or axis that show information of how `aes=variable` mapping is defined. However, those variable value expressions you see in the legend/axis are defined by `labels` (such as "Male", "Female"), whose corresponding variable values are defined in `breaks` (such as "m" and "f"); and `name` gives lengend/axis a title (such as "Gender").
</div>

## Time axis


```{r , eval=T, echo=F}
teachDS::img_centering("https://www.economist.com/img/b/1280/748/90/sites/default/files/20190112_WOC124.png", width="80%")
```

Generate basic plot:
```{r}
dataSet1 <- 
  data.frame(
    x=1979:2018
  )
set.seed(2038)
dataSet1$y <- sample(10:40, length(dataSet1$x), T)
ggplot1 <- list()

ggplot()+
  geom_step(
    data=dataSet1,
    mapping=
      aes(
        x=x,
        y=y
      )
  ) -> ggplot1$plot0
ggplot1$plot0
```

```{r}
# x
labels=c("1980", "1990", "2000", "2010")
breaks=c(1980, 1990, 2000, 2010)
```


***

Define x-axis labels:
```{r}
breaks = c(
  1979,
  seq(1985, 2015, by=5),
  2018
)
labels = c(
  "1979", "85", "90", "95", "2000", "05", "10", "15", "18"
)

ggplot1$plot0 +
  scale_x_continuous(
    breaks=breaks,
    labels=labels
  ) -> ggplot1$plot1

ggplot1$plot1
```

Remove scale ticks
```{r}
ggplot1$plot1 +
  theme(
    axis.ticks.length.x = unit(0,"mm")
  ) -> ggplot1$plot1

ggplot1$plot1
```

Add ticks:

  - major ticks: ticks that belong to constant time span labels
  
  - minor ticks: ticks that are not major ticks. 
  
```{r}
ticks <- list()
ticks$major <- seq(1980, 2015, by=5)
ticks$minor <- c(1979, 2018)

majorLength = 3 #input$length
minor_majorRatio = 0.7 #input$ratio

ggplot1$plot1 +
  geom_rug(
    mapping=aes(
      x=ticks$major
      ),
    outside=TRUE, # draw rug outside the plot panel
    size=0.5, #input$majorsize
    length=grid::unit(
      majorLength, 
      "mm"
    )
  ) +
  geom_rug(
    mapping=aes(
      x=ticks$minor
    ),
    outside = TRUE,
    size=0.5, #input$minorsize
    length=grid::unit(
      minor_majorRatio*majorLength,
      "mm"
    )
  )+
  coord_cartesian(clip="off") -> # allow drawing outside the plot panel 
  ggplot1$plot2

ggplot1$plot2
```


```{r}
ggplot1$plot2 +
  theme(
    axis.text.x = element_text(
      margin = margin(
        12 #input$margin
      ),
      size=16 #input$textSize
    ))
```

  * `margin(t=0, r=0, b=0, l=0, unit='pt')`

### Custom axis function

Pull out all the axis-related building blocks:
```{r}
{
  # scale_x 
    scale_x_continuous(
      breaks=breaks,
      labels=labels
    ) +
    theme(
      axis.ticks.length.x = unit(0,"mm"),
      axis.text.x = element_text(
        margin = margin(
          12 #input$margin
        ),
        size=16 #input$textSize
        )
      )+
    geom_rug(
      mapping=aes(
        x=ticks$major
        ),
      outside=TRUE, # draw rug outside the plot panel
      size=0.5, #input$majorsize
      length=grid::unit(
        majorLength, 
        "mm"
      )
    ) +
    geom_rug(
      mapping=aes(
        x=ticks$minor
      ),
      outside = TRUE,
      size=0.5, #input$minorsize
      length=grid::unit(
        minor_majorRatio*majorLength,
        "mm"
      )
    )+
    coord_cartesian(clip="off")
}
```

Build a function:
```{r axis_x_continuouse_custom}
axis_x_continuouse_custom <- function(
  breaks, labels,
  ticks_major, ticks_minor,
  ticks_major_length = 3, 
  minor_major_tickLength_ratio = 0.7,
  text_size = 16,
  text_top_margin = 12,
  major_tick_size = 0.5,
  minor_tick_size = 0.5
){
  list(
    scale_x_continuous(
      breaks=breaks,
      labels=labels
    ), 
    theme(
      axis.ticks.length.x = unit(0,"mm"),
      axis.text.x = element_text(
        margin = margin(
          text_top_margin #input$margin
        ),
        size=text_size #input$textSize
        )
      ), 
    geom_rug(
      mapping=aes(
        x=ticks_major
        ),
      outside=TRUE, # draw rug outside the plot panel
      size=major_tick_size, #input$majorsize
      length=grid::unit(
        ticks_major_length, 
        "mm"
      )
    ), 
    geom_rug(
      mapping=aes(
        x=ticks_minor
      ),
      outside = TRUE,
      size=minor_tick_size,
      length=grid::unit(
        minor_major_tickLength_ratio*ticks_major_length,
        "mm"
      )
    ), 
    coord_cartesian(clip="off")
    )
}

```

```{r}
breaks = c(
  1979,
  seq(1985, 2015, by=5),
  2018
)
labels = c(
  "1979", "85", "90", "95", "2000", "05", "10", "15", "18"
)
ticks_major <- seq(1980, 2015, by=5)
ticks_minor <- c(1979, 2018)

ggplot1$plot0 +
  axis_x_continuouse_custom(
    breaks=breaks, labels=labels, 
    ticks_major = ticks_major,
    ticks_minor = ticks_minor
  )
```

### Advanced function

  * Other input as default: when `labels` is omitted, use `breaks` value
  
  * When `ticks_minor` is omitted, remove minor geom_rug.

```{r}
axis_x_continuouse_custom <- function(
  breaks, labels = breaks,
  ticks_major, ticks_minor=NULL,
  ticks_major_length = 3, 
  minor_major_tickLength_ratio = 0.7,
  text_size = 16,
  text_top_margin = 12,
  major_tick_size = 0.5,
  minor_tick_size = 0.5
){
  list(
    scale_x_continuous(
      breaks=breaks,
      labels=labels
    ), 
    theme(
      axis.ticks.length.x = unit(0,"mm"),
      axis.text.x = element_text(
        margin = margin(
          text_top_margin #input$margin
        ),
        size=text_size #input$textSize
        )
      ), 
    geom_rug(
      data = data.frame(
        ticks_major=ticks_major
      ),
      mapping=aes(
        x=ticks_major
        ),
      outside=TRUE, # draw rug outside the plot panel
      size=major_tick_size, #input$majorsize
      length=grid::unit(
        ticks_major_length, 
        "mm"
      )
    ),
    if(!is.null(ticks_minor)){
      geom_rug(
        data = data.frame(
          ticks_minor=ticks_minor
        ),
        mapping=aes(
          x=ticks_minor
        ),
        outside = TRUE,
        size=minor_tick_size,
        length=grid::unit(
          minor_major_tickLength_ratio*ticks_major_length,
          "mm"
        )
      )
    } else {
      NULL
    }, 
    coord_cartesian(clip="off")
    )
}
```

```{r}
ggplot1$plot0 +
  axis_x_continuouse_custom(
    breaks=breaks,
    ticks_major = ticks_major
  )
```

***

There are many possible `scale_x`. We can build a `axis_x_custom` function that can take all possible `scale_x_zzz` as input, and return `axis_x_zzz_custom` function as an output.

  * Function that can take function as input is called **functional**.
  
  * Functionals that generate functions are called **function generators**.

```{r}
axis_x_custom <- function(scale_x){
  function(
    breaks, labels = breaks,
    ticks_major, ticks_minor=NULL,
    ticks_major_length = 3, 
    minor_major_tickLength_ratio = 0.7,
    text_size = 16,
    text_top_margin = 12,
    major_tick_size = 0.5,
    minor_tick_size = 0.5
  ){
    list(
      scale_x(
        breaks=breaks,
        labels=labels
      ), 
      theme(
        axis.ticks.length.x = unit(0,"mm"),
        axis.text.x = element_text(
          margin = margin(
            text_top_margin #input$margin
          ),
          size=text_size #input$textSize
          )
        ), 
      geom_rug(
        data = data.frame(
          ticks_major=ticks_major
        ),
        mapping=aes(
          x=ticks_major
          ),
        outside=TRUE, # draw rug outside the plot panel
        size=major_tick_size, #input$majorsize
        length=grid::unit(
          ticks_major_length, 
          "mm"
        )
      ),
      if(!is.null(ticks_minor)){
        geom_rug(
          data = data.frame(
            ticks_minor=ticks_minor
          ),
          mapping=aes(
            x=ticks_minor
          ),
          outside = TRUE,
          size=minor_tick_size,
          length=grid::unit(
            minor_major_tickLength_ratio*ticks_major_length,
            "mm"
          )
        )
      } else {
        NULL
      }, 
      coord_cartesian(clip="off")
      )
  }
}
```

***

```{r}
# generate axis_x_continuous_custom
axis_x_continuous_custom2 <-
  axis_x_custom(scale_x_continuous)
ggplot1$plot0 +
  axis_x_continuous_custom2(
    breaks=breaks,
    ticks_major = ticks_major
  )
ggplot1$plot0 +
  axis_x_continuous_custom2(
    breaks=breaks, labels=labels,
    ticks_major = ticks_major,
    ticks_minor = ticks_minor
  )
```


### Time period

When axis x is to represent a period:

```{r , eval=T, echo=F}
teachDS::img_centering("https://www.economist.com/img/b/1280/675/90/sites/default/files/20131221_FNC079.png")
```

Basic plot:
```{r}
dataSet2 <- data.frame(
  x=seq(from=lubridate::ymd("2013-01-01"),
    to=lubridate::ymd("2013-12-31"),
    by="1 day")
)

dataSet2$y <- {
  y <- c(100)
  set.seed(2033)
  shocks <- rnorm(length(dataSet2$x), sd=50)
  shocks
  for(t in 2:length(dataSet2$x)){
    y[[t]] <- 1*t + 0.6*y[[t-1]] + shocks[[t]]
  }
  y
}
```


7-days moving average:
```{r}
install.packages("zoo")
```


```{r}
library(dplyr)
dataSet2 %>%
  mutate(
    y_smooth=zoo::rollmean(y, 7, na.pad=TRUE, align="center")
  ) -> dataSet2
```

  * `zoo::rollmean(y, window_width, padding_na_to_maintain_length, window_center)`


```{r}
ggplot2 <- list()
ggplot2$plot1 <- {
  ggplot()+
    geom_line(
      data=dataSet2,
      mapping=aes(
        x=x,
        y=y_smooth
      )
    )
}

ggplot2$plot1
```

Add rug ticks and labels:
```{r}
axis_x_date_custom <- 
  axis_x_custom(scale_x_date)
```

```{r}
# breaks in middle of the month
#  set 15th of every month
breaks=seq(
  from=lubridate::ymd("20130115"),
  to=lubridate::ymd("20131215"),
  by="1 month"
)
labels=lubridate::month(
  breaks,
  label=TRUE # Use month name based on OS locale
)
labels

ticks_major=c(
  seq(
    from=lubridate::ymd("20130101"),
    to=lubridate::ymd("20131231"),
    by="1 month"
  ),
  lubridate::ymd("20131231"))
```

<div class="alert alert-info">
Locales in your operating system determine how month and weekday should be expressed:  
  * Windows: <https://docs.microsoft.com/en-us/openspecs/windows_protocols/ms-lcid/a9eac961-e77d-41a6-90a5-ce1a8b0cdb9c> 
  * Mac OS/linux:
```{r}
locales <- system("locale -a", intern = TRUE)
```

```{r}
lubridate::month(breaks, 
  locale = "zh_TW", 
  label = T # returned as month abbreviates
  )
lubridate::wday(breaks, 
  locale = "zh_TW", 
  label = T # returned as month abbreviates
  )
```

</div>

```{r}
ggplot2$plot1 +
  axis_x_date_custom(
    breaks=breaks, labels=labels,
    ticks_major=ticks_major,
    ticks_major_length = 2, #input$tickLength
    text_size = 14 #input$textSize
  )
```

### dot-dot-dot

  * `...` can pass any arguments not defined in function usage as one argument input that can be used in any place by calling `...` .

```{r}
axis_x_custom <- function(scale_x){
  function(
    breaks, labels = breaks,
    ticks_major, ticks_minor=NULL,
    ticks_major_length = 3, 
    minor_major_tickLength_ratio = 0.7,
    text_size = 16,
    text_top_margin = 12,
    major_tick_size = 0.5,
    minor_tick_size = 0.5, ...
  ){
    list(
      scale_x(
        breaks=breaks,
        labels=labels, ...
      ), 
      theme(
        axis.ticks.length.x = unit(0,"mm"),
        axis.text.x = element_text(
          margin = margin(
            text_top_margin #input$margin
          ),
          size=text_size #input$textSize
          )
        ), 
      geom_rug(
        data = data.frame(
          ticks_major=ticks_major
        ),
        mapping=aes(
          x=ticks_major
          ),
        outside=TRUE, # draw rug outside the plot panel
        size=major_tick_size, #input$majorsize
        length=grid::unit(
          ticks_major_length, 
          "mm"
        )
      ),
      if(!is.null(ticks_minor)){
        geom_rug(
          data = data.frame(
            ticks_minor=ticks_minor
          ),
          mapping=aes(
            x=ticks_minor
          ),
          outside = TRUE,
          size=minor_tick_size,
          length=grid::unit(
            minor_major_tickLength_ratio*ticks_major_length,
            "mm"
          )
        )
      } else {
        NULL
      }, 
      coord_cartesian(clip="off")
      )
  }
}

```

```{r}
axis_x_date_custom <-
  axis_x_custom(scale_x_date)

ggplot2$plot1 +
  axis_x_date_custom(
    breaks=breaks, labels=labels,
    ticks_major=ticks_major,
    ticks_major_length = 2, #input$tickLength
    text_size = 14, #input$textSize
    name="2013"
  )
```

***

Multiple rug tick lengths: 

```{r , eval=T, echo=F}
teachDS::img_centering("https://www.economist.com/img/b/1000/969/90/sites/default/files/images/print-edition/20140719_FNC335.png")
```

```{r}
dataSet3 <- data.frame(
  x=seq(from=lubridate::ymd("2011-01-01"),
    to=lubridate::ymd("2014-06-01"),
    by="1 month")
)

dataSet3$y <- {
  y <- c(100)
  set.seed(2033)
  shocks <- rnorm(length(dataSet3$x), sd=50)
  shocks
  for(t in 2:length(dataSet3$x)){
    y[[t]] <- 1*t + 0.6*y[[t-1]] + shocks[[t]]
  }
  y
}

dataSet3 %>%
  mutate(
    y_smooth = zoo::rollmean(y, 5, na.pad=TRUE, align="center")
  ) -> dataSet3 
```

```{r}
ggplot3 <- list()
ggplot()+
  geom_line(
    data=dataSet3,
    mapping=aes(
      x=x,
      y=y_smooth
    )
  ) -> ggplot3$plot1
ggplot3$plot1
```

```{r}
# breaks in middle of years, set at June, except 2014, set at March
breaks = c(
  seq(
    from=lubridate::ymd("20110601"),
    to=lubridate::ymd("20130601"),
    by="1 year"
  ),
  lubridate::ymd("20140301")
)
labels = c("2011", "12", "13", "14")
ticks_major=seq(
    from=lubridate::ymd("20110101"),
    to=lubridate::ymd("20140101"),
    by="1 year"
  )
ticks_minor=seq(
    from=lubridate::ymd("20110101"),
    to=lubridate::ymd("20140601"),
    by="1 month"
  )
```

```{r}
ggplot3$plot1 +
  axis_x_date_custom(
    breaks=breaks,
    labels=labels,
    ticks_major = ticks_major,
    ticks_minor = ticks_minor
  )
```

<div class="exercise">
How do you put ticks inside?
</div>


<div class="alert alert-info">
For date/time axis:  
  * always assign ticks to the beginning and the end of the sample period.  
  * when too many ticks are there,  
    * try to label few of them, like labeling every 5 years, every 6 months, etc. Or  
    * use period marking.  
  * when use period marking, put labels in the middle of the period.  
</div>

## XY axis

### Expansion and out-of-bound

Among `scale_x_...` and `scale_y_...`, there are two arguments that might be handy:

  * **expand**: define the padding around the limits of the plot. (圖形與xy軸的留白空間)
  
  * **oob** (out of bound): define what to do when observations are out of limits bound. (超出作圖範圍的資料點要怎麼處置)

```{r}
dataSet <- data.frame(
  x=rep(c("Jan", "Feb", "Mar"),2),
  y=c(10, 20, 30, 10, 25, 33),
  group=c(rep("Art", 3), rep("General", 3))
)
ggplot()+
  geom_col(
    data=dataSet,
    mapping=
      aes(
        x=x, y=y,
        fill=group
      ),
    position="dodge",
    width=0.8
  )+
  scale_y_continuous(
    expand = 
      expansion(0, #input$multiply
        0 #input$add 
        )
    )
```

  * `expand = c(lowerBoundStretch, upperBoundStretch)`:  
    * `lowerBoundStretch=c(a,b)`: `lowerBound - a*range -b`  
    * `upperBoundStretch=c(c,d)`: `upperBound + c*range +d`  
    
    * `expansion(m, n)` will set a=c=m, b=d=n.

### Secondary axis

```{r , eval=T, echo=F}
teachDS::img_centering("https://www.economist.com/img/b/1000/969/90/sites/default/files/images/print-edition/20140719_FNC335.png")
```

```{r}
dataSet4 <- data.frame(
  x=seq(
    from=lubridate::ymd("20110101"),
    to=lubridate::ymd("20140601"),
    by="1 month"
  )
)

dataSet4$unemployment <- 10+
  (12-10)*as.integer(dataSet4$x-dataSet4$x[[1]])/as.integer(dataSet4$x[[length(dataSet4$x)]]-dataSet4$x[[1]])
dataSet4$ind_prod <- 5 + (-5-5)*as.integer(dataSet4$x-dataSet4$x[[1]])/as.integer(dataSet4$x[[length(dataSet4$x)]]-dataSet4$x[[1]])
```

***

  * All layers are graphed based on the LHS y-axis.

We need:

  * A **break transfer function** to transfer LHS breaks to RHS breaks. (i.e. $0\rightarrow 11$, $5\rightarrow 12$)

```{r}
transfer_lr <- function(x) 11+(12-11)*x/5

breaks_lhs <- seq(-10,10, by=5)
breaks_rhs <- transfer_lr(breaks_lhs)
```
  
  * A **data series scale function** to scale RHS data series to LHS y-axis scale. (An inverse function of break transfer function)

```{r}
scale_rl <- function(y) 5*(y-11)

dataSet4$unemployment_scaled <- scale_rl(dataSet4$unemployment)
```

***

Basic plot:
```{r}
ggplot4 <- list()
ggplot(
  data=dataSet4,
  mapping=aes(x=x)
)+
  geom_col(
    mapping=aes(
      y=ind_prod
    ),
    fill="#04a2d0"
  ) -> ggplot4$plot1

ggplot4$plot1
```

Axis y without secondary axis:
```{r}
ggplot4$plot1 +
  scale_y_continuous(
    name="Industrial production",
    limits = c(-10, 10),
    breaks = seq(-10, 10, by=5)
  )
```

Axis y with secondary axis: 
```{r}
ggplot4$plot1 +
  scale_y_continuous(
    name="Industrial production",
    limits = c(-10, 10), # required. Since data range is less, without this, breaks setup will be truncated.
    breaks = seq(-10, 10, by=5),
    sec.axis = sec_axis(
      trans=transfer_lr, # map: first axis breaks to second axis breaks
      name="Unemployment",
      labels=c("//", as.character(10:13)) # secondary axis labels
    ),
    expand = c(0,0,0,
      1.5 #input$yupper
      )
  ) -> ggplot4$plot2

ggplot4$plot2
```


```{r}
ggplot4$plot2 +
  geom_line(
    aes(
      y=unemployment_scaled
    ),
    color="#77230f"
  ) -> ggplot4$plot3

ggplot4$plot3
```

<div class="alert alert-info">
`econDV2::geom_sec_axis` function turns any `geom` function into a secondary `geom` layer with equipped secondary y-axis on the right:
```{r}
# generate the geom for right y-axis
geom_line_right <- econDV2::geom_sec_axis(
  geom_line
)


ggplot4$y.axis$breaks_left = seq(-10, 10, by=5)
ggplot4$y.axis$breaks_right=9:13
ggplot4$y.axis$name_left = "Industrial production"
ggplot4$y.axis$name_right= "Unemployment"
ggplot4$y.axis$labels_right=c("//", as.character(10:13))

ggplot4$geom_line_right <- function(...){
    geom_line_right(
      mapping=aes(
        y=unemployment
      ),
      color="#77230f",
      breaks_left=ggplot4$y.axis$breaks_left,
      breaks_right=ggplot4$y.axis$breaks_right,
      name_left=ggplot4$y.axis$name_left, 
      name_right=ggplot4$y.axis$name_right,
      labels_right=ggplot4$y.axis$labels_right, ...
    )
}
ggplot4$plot1 +
  ggplot4$geom_line_right() -> ggplot4$plot3

ggplot4$plot3
```

Style axis-x:
```{r}
dataSet4$x |> class()
axis_x_date <- econDV2::axis_x_custom(scale_x_date)

axis_x_date |> args()

ggplot4$x.axis$breaks = seq(
  from=lubridate::ymd("20110601"),
  to=lubridate::ymd("20130601"),
  by="1 year"
)
ggplot4$x.axis$breaks = c(ggplot4$x.axis$breaks, lubridate::ymd("20140301"))
ggplot4$x.axis$labels = c("2011", "12", "13", "14")
ggplot4$x.axis$ticks_major= seq(
  from=lubridate::ymd("20110101"),
  to=lubridate::ymd("20140101"),
  by="1 year"
)
ggplot4$x.axis$ticks_minor=seq(
  from=lubridate::ymd("20110101"),
  to=lubridate::ymd("20140601"),
  by="1 month"
)

ggplot4$axis_x_date <- function(...){
  axis_x_date(
    breaks=ggplot4$x.axis$breaks,
    labels=ggplot4$x.axis$labels, 
    ticks_major = ggplot4$x.axis$ticks_major,
    ticks_minor=ggplot4$x.axis$ticks_minor,
    ...
  )
}

ggplot4$plot3 +
   ggplot4$axis_x_date() -> ggplot4$plot4

ggplot4$plot4
```


</div>

```{r}
size=8 #input$size_title
size_text=8 #input$size_text
vjust=1 #input$vjust
angle=0 #input$angel
margin_l=-50 #input$margin_l
margin_r=-50 #input$margin_r
ggplot4$plot1 +
  ggplot4$axis_x_date()+
  ggplot4$geom_line_right(
    expand=c(0,0,0, 
      0 #input$yupper
      )
  )+
  theme(
    axis.text.y.left=element_text(color="#04a2d0",
      size=size_text
      ),
    axis.text.y.right=element_text(color="#77230f",
      size=size_text
      ),
    axis.title.y.left = 
      element_text(
        color="#04a2d0", 
        size=size,
        vjust=vjust, 
        angle=angle, 
        margin=margin(
          r=margin_l
          )
        ),
    axis.title.y.right = 
      element_text(
        color="#77230f", 
        size=size,
        vjust=vjust, 
        angle=angle, 
        margin=margin(
          l=margin_r
          )
        ),
    axis.ticks.length.y=unit(0,"mm"),
    axis.line.y=element_blank(),
    panel.grid.major.y=element_line(color="#ececec")
  )
```


## Color/Fill

### Discrete


```{r , eval=T, echo=F}
teachDS::img_centering("https://www.economist.com/img/b/1280/755/90/sites/default/files/20200815_WOC191.png", width="80%")
```


Set up simulated data:
```{r}
dataSet3 <- data.frame(
  x=seq(
    from=lubridate::ymd("2020-01-03"),
    to=lubridate::ymd("2020-08-12"),
    by="1 day"
  )
)
countries <- c("Britain","France","Germany","Italy","Spain")
slopes <- c(0.1, 0.2, 0.3, 0.4, 0.5)
for(i in seq_along(slopes)){
  dataSet3[[countries[[i]]]] <-
    0 + slopes[[i]]*as.integer(dataSet3$x-dataSet3$x[[1]])
}
dataSet3 |>
  tidyr::pivot_longer(
    cols="Britain":"Spain",
    names_to = "country",
    values_to = "y"
  ) -> dataSet3
```

Initiate a base plot:
```{r}
ggplot3 <- list()
ggplot3$data <- dataSet3
ggplot3$plot1 <- function(){
  ggplot()+
    geom_line(
      data=ggplot3$data,
      mapping=aes(
        x=x,
        y=y,
        group=country,
        color=country
      )
    )
}

ggplot3$plot1()
```

<div class="alert alert-info">
Formulate the plotting process as a function has an advantage of data set substitutability. We can substitute `ggplot3$data` by:

```{r}
ggplot3$data <- new_data
ggplot3$plot1()
```

  * This is because function is lazily evaluated which will look for `ggplot3$data` when it is called.
  
  * The birth place of `ggplot3$plot1` function is global environment. Therefore, any update of `ggplot3$data` in global environment before a call of `ggplot3$plot1` will always be based on the updated value.

</div>

***

Prepare color scale:
```{r}
ggplot3$color$limits <- c("Britain", "France", "Germany", "Italy", "Spain")
ggplot3$color$values <- c("#984152", "#1e80ab", "#2ec1d2", "#af959f", "#e5b865")
ggplot3$color$labels <- c("英", "法", "德", "義", "西")
```

Formulate the `scale_color_manual` call as a function
```{r}
ggplot3$scale_color_manual <- function()
{
  limits = ggplot3$color$limits
  values = ggplot3$color$values
  
  # a call to scale_color_manual
  scale_color_manual(
    limits = limits,
    values = values,
    breaks = limits,
    labels = ggplot3$color$labels
  )
}
```


```{r}
ggplot3$plot1() +
  ggplot3$scale_color_manual()
```

<div class="alert alert-info">
`{...}` returns the visible value of the last executed line. 
</div>

```{r}
ggplot3$axis_x_date_monthlyPeriod <- function()
{
  dd <- data.frame(
    x = ggplot3$data$x
  )
  require(dplyr)
  dd %>%
    arrange(x) %>%
    mutate(
      year=lubridate::year(x),
      month=lubridate::month(x)
    ) %>%
    group_by(year,month) %>%
    distinct() %>%
    summarise(
      day1 = min(x),
      midlength = round(length(x)/2,0),
      breaks = day1 + lubridate::days(midlength),
      labels = lubridate::month(breaks, label=T)
    ) %>% ungroup() -> dx
  xrange <- range(dd$x)
  rug_x <- {
    c(xrange, dx$day1) |>
      unique() |>
      sort() 
  }
  
  list(
    geom_rug(
      mapping=aes(
        x=rug_x
      ),
      outside = T
    ),
    coord_cartesian(clip="off"),
    scale_x_date(
      breaks=dx$breaks,
      labels = dx$labels
    ),
    theme(
      axis.ticks.length.x = grid::unit(0,"mm")
    )
  )
}
```


```{r}
ggplot3$plot1() +
  ggplot3$scale_color_manual() +
  ggplot3$axis_x_date_monthlyPeriod()
```

### Continuous

```{r , eval=T, echo=F}
teachDS::img_centering("https://www.economist.com/img/b/1280/755/90/sites/default/files/20201031_WOC015.png", width="85%")
```

## Exercise

```{r , eval=T, echo=F}
teachDS::img_centering("https://www.economist.com/img/b/1000/2072/90/sites/default/files/images/print-edition/20211009_SRC217.png", width="80%")
```

```{r , eval=T, echo=F}
teachDS::img_centering("https://www.economist.com/img/b/1000/1153/90/sites/default/files/images/print-edition/20211030_FNC803.png", width="80%")
```
