# Categorical data

## Aesthetics: group

```{r}
ggplot() +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3),
      y=c(2, 3, 2),
    )
  ) +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3),
      y=c(5, 2, 6)
    )
  )

```

Use `group` aesthetic to combine 
  
  * multiple same geom layers

into one.

```{r}
ggplot() +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(2, 3, 2, 5, 2, 6),
      group=c("m", "m", "m", "f", "f", "f"), 
    )
  )

```

```{r}
ggplot() +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(2, 3, 2, 5, 2, 6),
      group=c("m", "m", "m", "f", "f", "f"),
      color=c("m", "m", "m", "f", "f", "f")
    )
  )

```

  * Any aesthetic differentiates group can replace group.
  
```{r}
ggplot() +
  geom_line(
    mapping=aes(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(2, 3, 2, 5, 2, 6),
      # group=c("m", "m", "m", "f", "f", "f"),
      color=c("m", "m", "m", "f", "f", "f")
    )
  )

```

<div class="alert alert-warning">

  * When there is no aesthetic mapping to differentiate groups, use `group` aesthetic mapping.
  
</div>

## Geom overlapping

When geom layers overlap, we can use

  * `alpha` aesthetic. 

If multiple geometries are created within the one `geom_` call (using grouping aesthetics), we can also set

  * position: "stack", "dodge" or "jitter" (some of them might not apply to certain `geom_`)

```{r , eval=T, echo=F}
teachDS::img_centering("https://clauswilke.com/dataviz/visualizing_proportions_files/figure-html/health-vs-age-1.png")
```

  * <https://clauswilke.com/dataviz/visualizing-proportions.html#fig:health-vs-age>
  
  * x is continuous, or discrete with many types
  
  * y the cumulative proportion

***

```{r}
ggplot() +
  geom_area(
    mapping=aes(
      x=c(1, 2, 3),
      y=c(0.2, 0.3, 0.2),
    )
  ) +
  geom_area(
    mapping=aes(
      x=c(1, 2, 3),
      y=c(0.4, 0.3, 0.52) + c(0.2, 0.3, 0.2) # the additive is for accumulative purpose
    ), 
    alpha=0.5
  )
```

## Position: stack

  * put y on top of the overlapping geom's y
  
  * create accumulative result.

```{r}
ggplot() +
  geom_area(
    mapping=aes(
      x=c(1, 2, 3, 
        1, 2, 3),
      y=c(0.2, 0.3, 0.2, 
        0.4, 0.3, 0.52),
      fill=c("m", "m", "m", 
        "f", "f", "f")
    ),
    position="stack" #input$position
  )
```

  * stack position is accumulative; no need to compute the accumulative value yourself.
  
  * the default position in `geom_area` is "stack". Therefore, you can omit position argument. 


***

```{r data_cat1}
data_cat1 <- data.frame(
      x=c(1, 2, 3, 1, 2, 3),
      y=c(0.2, 0.3, 0.2, 0.4, 0.4, 0.52),
      fill=c("m", "m", "m", "f", "f", "f")
)
```

```{r}
ggplot(
  data=data_cat1
) + 
  geom_area(
    mapping=aes(
      x=x,
      y=y,
      fill=fill
    )
  )
```

<div class="alert alert-warning">
When aesthetic mapping involves with unordered data, it will  

  * convert the data series into factor (unless the series is already a factor);
  
  * conduct the mapping according to the level sequence of the converted factor.

</div>

```{r}
data_cat1$fill |>
  factor() |>
  levels()
```


## Factor

When grouping aesthetics vary the look of geometries across different groups of data, it is crucial that users declare the mapped series with proper class.

  * `factor(data_series, levels)` parses `data_series` into a categorical data with expressing sequence defined by `levels`.
  
  * If omit `levels` the level sequence will be determined by the collateral sequence defined by your operating system.

```{r}
ggplot(
  data=data_cat1
) + 
  geom_area(
    mapping=aes(
      x=x,
      y=y,
      fill=factor(fill, levels=c("m", "f"))
    )
  )
```

  * Here we declare factor on-the-go.
  
We can also declare factor in the data frame first:
```{r}
data_cat1_copy <- data_cat1
data_cat1_copy$fill |>
  factor(levels=c("m", "f")) -> 
  data_cat1_copy$fill
```

  * `|>` is a R 4.0+ equipped operator, which makes:
```{r}
f(x, ....) # equivalent to
x |> f(...)
```
  

```{r}
ggplot(
  data=data_cat1_copy
) + 
  geom_area(
    mapping=aes(
      x=x,
      y=y,
      fill=fill
    )
  )

```

## Proportional data

```{r}
data_cat2_wide <- data.frame(
      x=c(1, 2, 3),
      y_a=c(0.2, 0.3, 0.2),
      y_b=c(0.4, 0.4, 0.52),
      y_c=c(0.4, 0.3, 0.28)
)

data_cat2_wide |> 
  tidyr::pivot_longer(
    cols=y_a:y_c,
    names_to = "fill",
    values_to= "y"
  ) ->
  data_cat2

View(data_cat2)
```


```{r}
ggplot(
  data=data_cat2
) + 
  geom_area(
    mapping=aes(
      x=x,
      y=y,
      fill=fill
    ),
    color="white"
  )
```

***

When x mapping series has limited cases and is discrete, a bar chart with position dodge is better.

```{r}
ggplot(
  data=data_cat2
) + 
  geom_col(
    mapping=aes(
      x=x,
      y=y,
      fill=fill
    ),
    color="white",
    width=0.8, #input$width
    size=0, #input$size
    position = "dodge" #input$position
  )
```

  * `width`: the width of the bar
  
  * `size`: the size of the stroke
  

***
Pie chart: 

  * not good for comparing proportion across more than one dimension 

```{r}
library(dplyr)
data_cat2 %>%
  filter(
    x==1
  ) -> 
  data_cat2_x1only
```

```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  )
```

```{r piechart}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  coord_polar(
    theta = "y"
  )
```

## Adding text 

adding text

```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=fill
    ),
    position = "stack"
  )
```

  * `geom_col` stack sequence is based on `fill` level sequence.
  
  * `geom_text` stack sequence is based on observation sequence.  

<div class="alert alert-info">

Grouping aesthetics determine the sequence of stacking. In `geom_col`, `fill` is the grouping aesthetic. To make `geom_text` stack labels in sequence as `fill` in `geom_col`, we can put `group=fill` in `geom_text` to create such a sequence. 

    
```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=fill,
      group=fill
    ),
    position = "stack"
  )
```

</div>

Change labels to represent the proportion values of `y`
```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=y, # use y to label now
      group=fill
    ),
    position = "stack"
  )
```

  * `position` argument also takes position functions.
  
  * When you know what type of position you want, you can use corresponding position function to fine tune the position.

```{r}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=y,
      group=fill
    ),
    position = position_stack(vjust=0.5)
  )
```

```{r piechart2}
ggplot(
  data=data_cat2_x1only
) + 
  geom_col(
    aes(
      x=x,
      y=y,
      fill=fill
    )
  ) +
  geom_text(
    aes(
      x=x,
      y=y,
      label=y,
      group=fill
    ),
    position = position_stack(vjust=0.5)
  ) +
  coord_polar(
    theta = "y"
  ) +
  theme_void()
```

***

When x-axis is also representing a categorical data:

```{r dd}
dy=0.03 # input$dy
ggplot(
  data=data_cat2
) + 
  geom_col(
    mapping=aes(
      x=x,
      y=y,
      fill=fill
    ),
    width=0.8, 
    position = "dodge" 
  )+
 geom_text(
    mapping=aes(
      x=x, 
      y=y,
      label=y,
      group=fill
    ),
    size = 8, #input$size
    position=position_dodge(
      width = 0.8 #input$width
    )
  )
```


```{r dd}
geom_text(
    mapping=aes(
      x=x,
      y=y-dy,
      group=fill,
      label=y
    ),
    size=8, #input$size
    position=position_dodge(width=
        0.8 #input$dodge
        )
  )
```

  * text position_dodge has the same width as `geom_col` to ensure the same dodging distance. 

## More on position

<https://ggplot2.tidyverse.org/reference/index.html#section-position-adjustment>

```{r}
ggplot(mpg, aes(x=class, y=hwy)) +
  geom_boxplot(colour = "grey50") +
  geom_point(alpha=0.2)
```

The lower and upper hinges correspond to the first and third quartiles (the 25th and 75th percentiles). 

## Coordination flip

```{r, eval=T, echo=F}
teachDS::img_centering("https://www.economist.com/img/b/1280/759/90/sites/default/files/20191130_WOC194.png")
```

```{r}
ggplot()+
  geom_col(
    mapping=
      aes(
        x=c("A", "B", "C"),
        y=c(56, 77, 92)
      )
  )+
  coord_flip()
```

***

Another common application of `coord_flip` is:

```{r , eval=T, echo=F}
teachDS::img_centering("img/us-presidential-election.jpg", width="100%")
```

```{r}
dx=4 #input$dx
h=0.5 #input$h
dt=0 #input$dt
ggplot()+
  geom_col(
    mapping=aes(
      x=c(1, 1),
      y=c(306, 232),
      fill=c("biden","trump")
    ),
    width=1
  )+
  geom_segment(
    mapping=aes(
      x=1-h,
      y=270,
      xend=1+h,
      yend=270
    )
  )+
  geom_text(
    mapping=aes(
      x=1+dt,
      y=270,
      label="270"
    ),
    size=8 #input$text
  )+
  xlim(1-dx, 1+dx)+ # make sure cover 0.5-1.5 so the bar width can be accomodate
  coord_flip()+
  theme_void()+
  theme(legend.position = "none")
```


## Summary

  * Grouping aesthetic separate a data frame into various subsample data frame and apply the `geom_` function to each one of them in the sequence determined by the mapping factor's levels sequence.
  
  * When `group` aesthetic and other aesthetic share the same mapping variable, `group` aesthetic can be ignored.
  
  * When deal with grouping variable, values of y from different groups at the same x can have position choice:  
    * "identity": respect ys as it is.  
    * "stack": stack ys according to grouping level sequence.
    * "dodge": respect ys as it is but move their x values left and right according to grouping level sequence. 

## Exercise

### 1{-}

```{r, echo=F, eval=T}
teachDS::img_centering("https://images.squarespace-cdn.com/content/v1/55b6a6dce4b089e11621d3ed/1586289284165-6Z2M2XKD645EJRCV3TPV/Area+chart+example.png?format=1500w")
```

### 2{-}

```{r, echo=F, eval=T}
teachDS::img_centering("https://www.economist.com/img/b/1000/1339/90/sites/default/files/images/print-edition/20211023_FBC728.png")
```

### 3{-}

```{r, echo=F, eval=T}
teachDS::img_centering("https://www.economist.com/img/b/1000/2753/90/sites/default/files/20211023_WWC082.png")
```


### 4{-}

```{r, echo=F, eval=T}
teachDS::img_centering("http://uc.udn.com.tw/photo/2020/01/12/1/7324036.png")
```

